---
title: "Likelihood-Based Estimation"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"

format: pptx

execute:
  echo: true
  warning: false
  message: false
---


```{r setup, echo=FALSE}
library(tidyverse)
library(broom)
library(scales)

set.seed(104)

biomarker_df <- tibble(
  id = 1:180,
  treatment = rep(c("Placebo", "Low dose", "High dose"), each = 60),
  ifng = c(
    rnorm(60, mean = 11.5, sd = 4.2),
    rnorm(60, mean = 15.3, sd = 4.8),
    rnorm(60, mean = 19.8, sd = 5.1)
  ),
  response = c(
    rbinom(60, 1, 0.28),
    rbinom(60, 1, 0.45),
    rbinom(60, 1, 0.62)
  ),
  infections = c(
    rpois(60, lambda = 1.4),
    rpois(60, lambda = 1.1),
    rpois(60, lambda = 0.8)
  ),
  age = round(rnorm(180, mean = 61, sd = 10)),
  bmi = round(rnorm(180, mean = 29, sd = 4.5), 1)
)
```

### Lecture Objectives

::: {style="font-size:0.6em"}
-   define likelihood as a general estimation principle
-   sample summaries estimate population quantities using only observed group-level data \[i.e.: sample means, proportions\]
-   discuss how model-based estimation use statistical models to estimate relationships between variables \[i.e.: regression\]
-   high level interpretation of linear and logistic regression output at an introductory level
-   connect coefficients to clinically meaningful effect measures
:::

------------------------------------------------------------------------

### Transition to Model-Based Estimation

::: {style="font-size:0.6em"}
Previous lectures focused on estimators that describe what is observed in the sample:

-   sample means → estimate population means
-   group differences → estimate contrasts between populations

Statistical summaries are limited when addressing questions such as:

-   how outcomes change across multiple predictors
-   how to handle outcomes with different structures \[continuous, binary, counts\]

Statistical models provide a framework to express how an outcome depends on predictors through unknown parameters that must be estimated
:::

------------------------------------------------------------------------

### Transition to Model-Based Estimation

::: {style="font-size:0.6em"}
**Goal**: estimate unknown population quantities using observed data

**Direct estimators**

-   computed from observed data
-   summarize outcomes within groups
-   do not account for additional variables

**Model-based estimators**

-   specify a relationship between outcome and predictors
-   estimate unknown parameters that define that relationship
-   allow adjustment for multiple variables and more complex structures
:::

------------------------------------------------------------------------

### Model Based Estimation

::: {style="font-size:0.6em"}
Model-based estimation:

approach to estimation that uses a statistical model to describe how the outcome is generated and to estimate unknown parameters

Key components:

-   **outcome \[Y\]**: variable being studied
-   **predictors \[X\]**: variables used to explain variation in the outcome
-   **parameters \[β\]**: unknown quantities describing the relationship between X and Y
-   **model**: mathematical expression linking Y to X through β

Estimation here means: use observed data to estimate the unknown parameters that define how the outcome depends on the predictors
:::

------------------------------------------------------------------------

### Likelihood

::: {style="font-size:0.6em"}
**likelihood** : function of the unknown parameters \[i.e.: probability of an event\] that describes the probability of the observed data under a specified statistical model

-   function that expresses how plausible different values of the model parameters are given the observed data

-   in estimation the goal is to learn about unknown population parameters using observed data

    -   data Y and predictors X are fixed once observed
    -   parameters β are unknown and must be **estimated**
:::

------------------------------------------------------------------------

### Likelihood

::: {style="font-size:0.6em"}
Likelihood can be expressed as follows:

$$
L(\beta) = P(Y \mid X, \beta)
$$

formula states that the likelihood evaluates how plausible the observed outcomes are under a model with parameters $\beta$ given the observed data

-   $L(\beta)$ : likelihood, viewed as a function of the unknown parameters
-   $Y$ : set of observed outcomes
-   $X$ : set of observed predictors or covariates
-   $\beta$ : collection of unknown model parameters

Likelihood estimate quality:

-   `good estimate` : observed data likely under the assumed model
-   `poor estimate` : parameter values under which the observed data would be unlikely
:::

------------------------------------------------------------------------

### Maximum Likelihood Estimation

::: {style="font-size:0.6em"}
**maximum likelihood estimation (MLE)** : method for estimating unknown model parameters by selecting the values that maximize the likelihood function

-   chooses the parameter values under which the observed data are most plausible

MLE Formula:

$$
\hat{\beta}=argmax_{\beta}L(β)
$$

-   $\hat{\beta}$ : estimated parameter values\
-   $L(\beta)$ : likelihood function\
-   $\arg\max$ : values of $\beta$ that maximize the likelihood
:::

------------------------------------------------------------------------

### Maximum Likelihood Estimation

::: {style="font-size:0.6em"}
**MLE involves**:

-   evaluating many possible parameter values
-   selecting the value that best aligns the model with the observed data
-   resulting estimate is the one under which the observed dataset is most consistent with the model

**Linking Estimation, Likelihood, and MLE**

-   `estimation` : learn about unknown population parameters using observed data
-   `likelihood` : approach for evaluating how well different parameter values explain the observed data
-   `MLE` : method that selects the parameter values that maximize the likelihood function $\rightarrow$ yields the estimate most consistent with the observed data

MLE provides a general framework for estimation across models such as:

-   linear regression
-   logistic regression
:::

------------------------------------------------------------------------

### Linear Regression as Model-Based Estimation

::: {style="font-size:0.6em"}
**linear regression model**: statistical model that represents the expected value of a continuous outcome as a linear function of predictor variables with unknown parameters estimated from observed data

-   linear regression - parameters β define the relationship between predictors (X) and outcome (Y)
-   parameters are estimated from the data using maximum likelihood

Model specification \[simple linear regression\]

-   conditional mean of the outcome is a linear function of the predictor

$$
E(Y∣X)=β_0+β_1X
$$

Where:

-   $E(Y \mid X)$ : xpected value of the outcome given the predictors $X$

-   $\beta_0$ : intercept → expected outcome when the predictor equals zero or when all predictors are at their reference values $X = 0$

-   $\beta_1$ : slope → expected change in the outcome for a one-unit change in the predictor $X$

-   $X$ : predictor
:::

------------------------------------------------------------------------

### Linear Regression as Model-Based Estimation

::: {style="font-size:0.6em"}
-   linear model defines a probability distribution for the observed data

-   likelihood evaluates how plausible different parameter values $(\beta, \sigma^2)$

-   unknown parameters include:

    -   $\beta_0, \beta_1$ : describes how the expected outcome depends on $X$
    -   $\sigma^2$ : describes how much outcomes vary around the mean

-   likelihood function:

    -   evaluates how plausible the observed data are for different values of $(\beta, \sigma^2)$
    -   data is fixed and varies the parameter values
:::

------------------------------------------------------------------------

### Linear Regression: Example

::: {style="font-size:0.6em"}
**Background**: In many clinical and translational studies, immune system activity is measured using circulating biomarkers. One commonly studied marker is interferon-gamma (IFN-γ), a cytokine that plays a central role in immune activation, inflammation, and host defense against infections. IFN-γ is often used to evaluate whether a therapeutic intervention can modify an immune response.

**Aim**: Consider a study designed to evaluate a new therapy compared to a placebo control, where the goal is to understand both biological response and clinical outcomes. The scientific motivation is grounded in questions such as:

-   How does IFN-γ differ across treatment groups?
-   What is the magnitude of those differences?
:::

------------------------------------------------------------------------

### Linear Regression: Example

::: {style="font-size:0.6em"}
**Description of the Dataset**

```{r}
head(biomarker_df, 3)
```

Each row in the dataset represents one patient enrolled in the study. The variables capture both biological measurements and clinical outcomes:

-   **id**: unique patient identifier
-   **treatment**: treatment group assignment \[placebo or low dose therapy\]
-   **ifng**: continuous measurement of IFN-γ concentration \[immune biomarker\]
-   **response**: binary clinical outcome \[1 = responded to treatment, 0 = no response\]
-   **infections**: count of infection events during follow-up
-   **age**: patient age in years
-   **bmi**: body mass index
:::

------------------------------------------------------------------------

### Linear Regression for IFN-gamma

::: {style="font-size:0.6em"}
Expected \[average\] IFN-γ level for each treatment group

$$
E(\text{IFNy∣treatment})=β_0+β_1⋅treatment
$$

Where:

$\beta_0$ : mean IFN-γ in the reference group \[high-dose group\]\
$\beta_1$ : difference in mean IFN-γ between low dose treatment and placebo

-   intercept \[$\beta_0$\] : represents the estimated mean IFN-γ in the high-dose group
-   treatment coefficient \[$\beta_1$\] : represents the estimated difference in mean IFN-γ between groups
:::

------------------------------------------------------------------------

### Linear Regression for IFN-gamma

::: {style="font-size:0.6em"}
```{r, echo=FALSE}
fit_lm <- lm(ifng ~ treatment, data = biomarker_df)
summary(fit_lm)
```

-   `lm()` function in base R fits linear model
-   model estimates the average IFN-gamma level as a function of treatment group
:::

------------------------------------------------------------------------

### Linear Regression for IFN-gamma

::: {style="font-size:0.5em"}
```{r, echo=FALSE}
fit_lm <- lm(ifng ~ treatment, data = biomarker_df)
summary(fit_lm)
```

**Interpretation of Treatment Effects**

Each coefficient represents a difference in mean IFN-γ relative to the reference group \[high dose treatment group\]:

Intercept = 19.55

-   estimated mean IFN-γ in the reference group \[high-dose treatment\]
:::

------------------------------------------------------------------------

### Linear Regression for IFN-gamma

::: {style="font-size:0.5em"}
```{r, echo=FALSE}
fit_lm <- lm(ifng ~ treatment, data = biomarker_df)
summary(fit_lm)
```

**Interpretation of Treatment Effects**

Low dose: −2.98

-   IFN-γ is about 3 units lower than the high-dose group
-   p = 0.000294 → strong evidence this difference is real

Placebo: −8.11

-   IFN-γ is about 8 units lower than the high-dose group
-   p \< 2e−16 → strong evidence of a difference between groups
:::

------------------------------------------------------------------------

### Linear Regression for IFN-gamma

::: {style="font-size:0.5em"}
```{r, echo=FALSE}
fit_lm <- lm(ifng ~ treatment, data = biomarker_df)
summary(fit_lm)
```

**Interpretation of Treatment Effects**

Suggested dose–response pattern:

Highest IFN-γ response: high-dose treatment Intermediate IFN-γ response: low dose Lowest IFN-γ response: placebo

-   therapy increases immune activation

-   magnitude of increase is dose-dependent

-   intervention is mechanistically active

-   linear model estimates group-specific mean IFN-γ levels

-   quantifies differences between treatment groups
:::

------------------------------------------------------------------------

### Visualizing Fitted Means from the Linear Model

```{r, echo=FALSE}

#| fig-width: 4 
#| fig-height: 3


lm_pred_df <- tibble(
  treatment = c("Placebo", "Low dose", "High dose")
)

lm_pred_df <- lm_pred_df |>
  mutate(
    predicted_ifng = predict(fit_lm, newdata = lm_pred_df)
  )

ggplot(lm_pred_df, aes(x = treatment, y = predicted_ifng)) +
  geom_col() +
  labs(
    title = "Linear Model Estimated Mean IFN-gamma \n by Treatment",
    x = "Treatment group",
    y = "Estimated mean IFN-gamma"
  ) +
  theme_bw()
```

::: {style="font-size:0.5em"}
**Interpretation**

-   bar plot showing estimated mean IFN-gamma levels for each treatment group based our linear model

-   High dose: estimated mean ≈ 19.6 pg/mL \[model intercept\]

-   Low dose: estimated group mean difference ≈ 16.6 pg/mL \[19.55−2.98=16.57\]

-   Placebo: estimated group mean difference ≈ 11.5 pg/mL \[19.55−8.11=11.45\]
:::

------------------------------------------------------------------------

### Logistic Regression

::: {style="font-size:0.6em"}
**logistic regression** : models how the probability of an event changes across predictors for binary outcomes

-   estimates and interprets how predictors influence the probability of a binary outcome

-   models the probability that an event occurs as a function of predictors

-   model estimates differences in event probability expressed through odds

-   model provides estimated probabilities for each group and uncertainty through standard errors and confidence intervals
:::

------------------------------------------------------------------------

### Logistic Regression

::: {style="font-size:0.6em"}
Model expresses the log-odds of the event as a linear function of the predictors:

$$
\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X
$$

-   formula states that the logarithm of the event odds is modeled as a linear function of the predictors

-   $p$ : probability of the event for a given value of $X$

    e.x.: probability that a patient responds to treatment \[responder vs. non-responder\]

-   $\frac{p}{1-p}$ : compares the probability the event occurs to the probability it does not

    how likely the event is to occur relative to non-occurrence

-   $\log\left(\frac{p}{1-p}\right)$ : log-odds \[logit\]

-   $\beta_0$ : intercept

-   $\beta_1$ : coefficient associated with the predictor

-   $X$ : predictor
:::

------------------------------------------------------------------------

### Logistic Regression for Response

::: {style="font-size:0.6em"}
Research Question:

> How does treatment affect the probability of clinical response?

Where:

-   reference group is the high-dose treatment group
-   model estimates:
    -   baseline log-odds \[high dose\]
    -   differences in log-odds for low dose and placebo
:::

------------------------------------------------------------------------

### Logistic Regression for Response

::: {style="font-size:0.6em"}
```{r, echo=FALSE}
fit_logit <- glm(response ~ treatment, data = biomarker_df, family = binomial())
summary(fit_logit)
```

**Interpretation**

Intercept \[Baseline Probability\]

-   Intercept = 0.928 - log-odds of response in the high-dose group

Treatment Effects

-   treatment coefficients : estimated difference in log-odds of response relative to high dose group
:::

------------------------------------------------------------------------

### Logistic Regression for Response

::: {style="font-size:0.6em"}
**Interpretation**

-   model estimates a difference in log-odds
-   exponentiating converts this into an odds ratio
-   odds ratio tells you how many times larger or smaller the odds are in one group compared to a reference group

Low Dose vs High Dose

-   estimate = −1.196
-   odds ratio = $e^{-1.196}$ ≈0.30
-   patients on low dose have about 70% lower odds of response compared to high dose
-   p = 0.002 : statistically significant association
:::

------------------------------------------------------------------------

### Logistic Regression for Response

::: {style="font-size:0.6em"}
**Interpretation**

Placebo vs High Dose

-   estimate = −2.027
-   odds ratio = $e^{-2.207}$ ≈0.13
-   placebo patients have about 87% lower odds of response compared to high dose
-   p \< 0.001 : statistically significant association

Note:

-   odds ratio = 1 : equal odds relative to the reference group
-   odds ratio \> 1 : higher odds of response than the reference group
-   odds ratio \< 1 : lower odds of response than the reference group
:::

------------------------------------------------------------------------

### Logistic Regression for Response

::: {style="font-size:0.6em"}
Note:

-   odds ratio = 1 : equal odds relative to the reference group
-   odds ratio \> 1 : higher odds of response than the reference group
-   odds ratio \< 1 : lower odds of response than the reference group
:::

------------------------------------------------------------------------

### Review: Model-Based Estimation

::: {style="font-size:0.6em"}
model-based estimates can:

-   compare multiple groups at once
-   adjust for additional predictors
-   provide fitted values on appropriate scales
-   estimate interpretable contrasts such as mean differences, odds ratios, and rate ratios
:::

------------------------------------------------------------------------

### Summary

::: {style="font-size:0.6em"}
-   likelihood evaluates how plausible the observed data are under different parameter values
-   maximum likelihood estimation selects the parameter values that make the observed data most plausible under the model
-   probability models and likelihood to extend estimation from summaries to structured predictor-outcome relationships
-   linear regression models conditional means for continuous outcomes with coefficients representing how the expected population average changes with predictors.
-   logistic regression models log-odds for binary outcomes and yields odds-ratio interpretations after exponentiation
-   exponentiated coefficients give odds ratios $\rightarrow$ quantifying relative changes in outcome likelihood
-   later regression lectures will build on concepts introduced in the current lecture
:::

------------------------------------------------------------------------
