---
title: "Bootstrap Estimation for Mean and Median Biomarker Values"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    chalkboard: true
    smaller: false
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r setup, echo=FALSE}
library(tidyverse)
library(broom)
library(scales)

set.seed(104)

biomarker_df <- tibble(
  id = 1:180,
  treatment = rep(c("Placebo", "Low dose", "High dose"), each = 60),
  ifng = c(
    rlnorm(60, meanlog = log(11.5), sdlog = 0.35),
    rlnorm(60, meanlog = log(15.3), sdlog = 0.38),
    rlnorm(60, meanlog = log(19.8), sdlog = 0.42)
  ),
  response = c(
    rbinom(60, 1, 0.28),
    rbinom(60, 1, 0.45),
    rbinom(60, 1, 0.62)
  ),
  infections = c(
    rpois(60, lambda = 1.4),
    rpois(60, lambda = 1.1),
    rpois(60, lambda = 0.8)
  ),
  age = round(rnorm(180, mean = 61, sd = 10)),
  bmi = round(rnorm(180, mean = 29, sd = 4.5), 1)
) |>
  mutate(
    treatment = factor(treatment, levels = c("Placebo", "Low dose", "High dose")),
    log_ifng = log(ifng)
  )

# add a small number of extreme values to motivate robust estimation
biomarker_extreme_df <- biomarker_df |>
  mutate(
    ifng_extreme = if_else(id %in% c(8, 77, 146), ifng * 4.5, ifng),
    log_ifng_extreme = log(ifng_extreme)
  )
```

### Review

::: {style="font-size:0.63em"}
Previous lectures established what is typically estimated in statistical analysis—defining population quantities and constructing estimators to compare groups \[i.e.: mean differences, risk differences, odds ratios\]

Current lecture focuses on the following:

> Given an estimator and a single observed dataset, how do we quantify its sampling variability when formulas rely on assumptions that may not reflect the observed data?

Bootstrap resampling provides a data-driven way to approximate the sampling distribution of an estimator
:::

------------------------------------------------------------------------

### Lecture Overview

::: {style="font-size:0.63em"}
-   define bootstrap resampling as a method for estimating sampling variability

-   interpret a bootstrap standard error and bootstrap confidence interval

-   explain why robust estimators are useful for skewed or outlier-prone biomedical measurements

-   compare the mean and median as estimators of a typical biomarker value

-   choose an estimation strategy based on the estimand, data structure, assumptions, and scientific question
:::

------------------------------------------------------------------------

### Bootstrap: Definition

::: {style="font-size:0.64em"}
`bootstrap`: an approach that approximates the sampling distribution of an estimator by repeatedly resampling from the observed data with replacement

Bootstrap Approach:

-   treats the observed dataset as a proxy for the population
-   generates many “new” samples by sampling with replacement
-   computes the estimator in each resample
-   uses the resulting distribution to quantify variability
:::

------------------------------------------------------------------------

### Bootstrap: Definition

::: {style="font-size:0.64em"}
Goals within statistical estimation include:

-   how much an estimate varies across samples
-   how precise the estimate is
-   what values are plausible for the population

Approaches rely on:

-   known sampling distributions
-   analytic standard error formulas
-   assumptions about the data \[i.e.: normality\]

However in certain cases: assumptions do not align with observed data

-   closed-form methods for estimating uncertainty rely on assumptions about the data are true in order to estimate sample variability \[i.e.: normality and simple variance structure\]

-   bootstrap methods estimate uncertainty by repeatedly resampling the data and observing how the estimate changes
:::

------------------------------------------------------------------------

### Overview: Bootstrap Procedure

::: {style="font-size:0.64em"}
Given observed data of size n:

**1. Resample with replacement**

-   draw a sample of size n from the original data

**2. Compute the estimator** - i.e.: mean, median, risk difference

**3. Repeat many times** - i.e.: 1,000+ resamples

**4. Form the bootstrap distribution** - collection of all computed estimates

Resulting distribution approximates the sampling distribution of the estimator
:::

------------------------------------------------------------------------

### Overview: Bootstrap Procedure

::: {style="font-size:0.64em"}
From the bootstrap distribution, we can estimate:

1.  **Standard error**: measures variability across resamples

2.  **Confidence intervals**: provides a plausible range for the population parameter

3.  **Shape of the sampling distribution** :

-   symmetry vs skewness
-   sensitivity to outliers
-   stability of the estimator
:::

------------------------------------------------------------------------

### Example: Estimating Typical Tumor Mutation Burden

::: {style="font-size:0.68em"}
**Background**

Investigators are studying tumor genomic profiles from patients with solid tumors.

The quantitative genomic measure of interest is tumor mutation burden, or TMB.

Tumor mutation burden is the number of somatic mutations observed per megabase of sequenced tumor DNA.

TMB provides a summary of how many mutations are present in a tumor genome and is commonly used to describe variation in tumor genomic profiles across patients.

In this study, TMB is measured from tumor sequencing data to summarize the mutational profile of the patient cohort.

:::

------------------------------------------------------------------------

### Example: Estimating Typical Tumor Mutation Burden

::: {style="font-size:0.68em"}
**Background**

TMB values are often right-skewed:

-   many tumors have low-to-moderate mutation burden
-   fewer tumors have very high mutation burden
-   extreme values can pull the mean upward

Investigators want to estimate:

> What is a typical tumor mutation burden among patients in this sequencing cohort?
:::

-------------

### Simulated Tumor Mutation Burden Dataset

::: {style="font-size:0.68em"}

```{r, echo=FALSE}
library(tidyverse)

set.seed(88)

n <- 220

tmb_df <- tibble(
  patient_id = 1:n,
  
  tumor_type = sample(
    c("Lung", "Colorectal", "Melanoma", "Breast"),
    size = n,
    replace = TRUE,
    prob = c(0.30, 0.28, 0.22, 0.20)
  ),
  
  tumor_stage = sample(
    c("I", "II", "III", "IV"),
    size = n,
    replace = TRUE,
    prob = c(0.18, 0.27, 0.31, 0.24)
  ),
  
  age = round(
    rnorm(
      n,
      mean = 62,
      sd = 10
    )
  ),
  
  tmb = round(
    rgamma(
      n,
      shape = 2.4,
      scale = 3.1
    ),
    2
  ),
  
  high_tmb_status = rbinom(
    n,
    size = 1,
    prob = 0.34
  )
)

head(tmb_df, 5)
```

**Dataset Overview**

each row represents one tumor sample from one patient

-   `tmb` : tumor mutation burden measured as mutations per megabase
-   `tumor_type` : cancer type represented in the sequencing cohort
-   `tumor_stage` : clinical stage at sampling
-   `age` : patient age at sequencing
-   `high_tmb_status` : binary indicator for whether TMB is at least 10 mutations per megabase

:::

------------------------------------------------------------------------

### Scientific Question and Estimands

::: {style="font-size:0.52em"}
Primary scientific question:

> What is a typical tumor mutation burden among patients with solid tumors in the target sequencing population?

Population quantities of interest:

$θ_{mean}$ = population mean tumor mutation burden 

$θ_{median}$ = population median tumor mutation burden

Corresponding estimators:

$\bar{x}$ = sample mean tumor mutation burden 

$\bar{x}$ = sample median tumor mutation burden

Bootstrap methods will be used to estimate:

-   standard error of the mean TMB
-   standard error of the median TMB
-   stability of each estimator across resampled datasets
:::

------------------------------------------------------------------------

### Visualize the TMB Distribution

::: {style="font-size:0.55em"}
```{r, echo=FALSE}
plot <- ggplot(
  tmb_df,
  aes(x = tmb)
) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Tumor Mutation Burden",
    subtitle = "Right-skewed genomic measurements are common in \n sequencing data",
    x = "Tumor mutation burden (mutations per megabase)",
    y = "Number of tumors"
  ) +
  theme_bw()

plot
```

**Interpretation of the Distribution**

-   histogram shows the distribution of tumor mutation burden across sequenced tumors
-   distribution is right-skewed positivelyskewed
-   due to skewness the mean is likely larger than the median
:::

------------------------------------------------------------------------

### Visualize the TMB Distribution

::: {style="font-size:0.5em"}
```{r, echo=FALSE}
plot <- ggplot(
  tmb_df,
  aes(x = tmb)
) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Tumor Mutation Burden",
    subtitle = "Right-skewed genomic measurements are common in \n sequencing data",
    x = "Tumor mutation burden (mutations per megabase)",
    y = "Number of tumors"
  ) +
  theme_bw()

plot
```

**Interpretation of the Distribution**

Note: mean may generate an overestimation of the typical tumor mutation burden in this sample because of extreme values

-   most tumors contain relatively low-to-moderate TMB values
-   smaller number of tumors contain substantially elevated mutation burden
-   long right tail indicates genomic heterogeneity across tumors
:::

------------------------------------------------------------------------

### Bootstrap Appraoch

::: {style="font-size:0.68em"}
`bootstrap resampling`: approximates repeated sampling using the observed data

Each bootstrap iteration executes the following:

**1. resample TMB values** with replacement

-   after a value is selected, it is returned to the dataset and can be selected again
-   allowing the bootstrap sample to:
    -   maintain same size as the original dataset
    -   note that each resample may include some participants more than once and may omit some participants

**2. compute the estimator on that resampled dataset**

**3. store the result**

**4. repeat many times**

After many iterations the collection of estimates approximates the sampling distribution of the estimator
:::

------------------------------------------------------------------------

### Extract the Variable

::: {style="font-size:0.6em"}
```{r}
tmb_values <- tmb_df$tmb

head(tmb_values)
```

**Notes**

-   `tmb_df$tmb` extracts the tumor mutation burden column
-   bootstrap resampling operates on the observed TMB measurements
-   vector contains one TMB value per tumor sample
:::

------------------------------------------------------------------------

### Generate One Bootstrap Sample

::: {style="font-size:0.6em"}
```{r}
set.seed(100)

bootstrap_sample <- sample(
  x = tmb_values,
  size = length(tmb_values),
  replace = TRUE
)

head(bootstrap_sample)
```

**Notes**

-   `sample()` draws observations from the original TMB values

-   `x = tmb_values`

    -   variable that stores values being resampled

-   `size = length(tmb_values)`

    -   bootstrap sample has the same size as the original dataset

-   `replace = TRUE`

    -   observations can appear more than once
:::

------------------------------------------------------------------------

### Compute an Estimator

::: {style="font-size:0.6em"}
```{r}
mean(bootstrap_sample)
median(bootstrap_sample)
```

-   represents estimators \[sample mean and median TMB\] from a single bootstrap sample

After generating the resample:

-   compute the estimator of interest
-   store the result
-   repeat many times
:::

------------------------------------------------------------------------

### Bootstrap for the Mean TMB

::: {style="font-size:0.5em"}
```{r}
set.seed(222)

boot_means <- replicate(2000, {

  x_star <- sample(
    tmb_df$tmb,
    size = nrow(tmb_df),
    replace = TRUE
  )

  mean(x_star)

})

head(boot_means)
```

**Notes**

-   `replicate(2000, { ... })` : repeats the bootstrap procedure 2000 times

Within each iteration:

-   generate a bootstrap resample of TMB values
-   compute the sample mean TMB
-   return the computed value

Result:

-   `boot_means` contains 2000 bootstrap estimates
-   collection approximates the sampling distribution of the mean TMB
:::

------------------------------------------------------------------------

### Convert Estimates into a Distribution

::: {style="font-size:0.5em"}
```{r, echo=FALSE}
boot_mean_df <- tibble(
  mean_tmb = boot_means
)

plot <- ggplot(boot_mean_df, aes(x = mean_tmb)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Bootstrap Distribution of the Mean TMB",
    x = "Bootstrap Mean TMB",
    y = "Count"
  ) +
  theme_bw()

plot
```

**Interpretation**

-   each bar represents bootstrap-estimated mean TMB values
-   histogram approximates the sampling distribution of the mean
-   spread reflects sampling variability

:::

------------------------------------------------------------------------

### Convert Estimates into a Distribution

::: {style="font-size:0.5em"}
```{r, echo=FALSE}
boot_mean_df <- tibble(
  mean_tmb = boot_means
)

plot <- ggplot(boot_mean_df, aes(x = mean_tmb)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Bootstrap Distribution of the Mean TMB",
    x = "Bootstrap Mean TMB",
    y = "Count"
  ) +
  theme_bw()

plot
```

**Interpretation**

Narrow distribution:

-   mean TMB is relatively stable across resamples

Wide distribution:

-   mean TMB varies more across resampled datasets
:::

---------

### Bootstrap for the Median TMB

::: {style="font-size:0.6em"}
```{r}
set.seed(333)

boot_medians <- replicate(2000, {

  x_star <- sample(
    tmb_df$tmb,
    size = nrow(tmb_df),
    replace = TRUE
  )

  median(x_star)

})

head(boot_medians)
```

**Notes**

-   code repeats the same bootstrap procedure for median
-   boot_medians contains 2000 bootstrap median TMB estimates
-   distribution approximates the sampling distribution of the median TMB
:::

------------------------------------------------------------------------

### Convert Estimates into a Distribution

::: {style="font-size:0.6em"}
```{r, echo=FALSE}
boot_median_df <- tibble(
  median_tmb = boot_medians
)

plot <- ggplot(boot_median_df, aes(x = median_tmb)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Bootstrap Distribution of the Median TMB",
    x = "Bootstrap Median TMB",
    y = "Count"
  ) +
  theme_bw()

plot
```

**Interpretation**

-   each bar represents bootstrap-estimated median TMB values
-   histogram approximates the sampling distribution of the median
-   spread reflects how much the median changes across resamples
-   median is less affected by extreme high TMB values
-   bootstrap helps quantify uncertainty for this robust estimator
:::

------------------------------------------------------------------------

### Estimating the Bootstrap Standard Error

::: {style="font-size:0.4em"}
```{r}
boot_mean_se <- sd(boot_means)
boot_median_se <- sd(boot_medians)

tibble(
  estimator = c("Mean TMB", "Median TMB"),
  bootstrap_se = c(boot_mean_se, boot_median_se)
)
```

-   bootstrap standard error is computed directly from the bootstrap estimates
-   measures how much the estimator varies across repeated resamples

$$
SE_{\mathrm{boot}}
=
\sqrt{
\frac{1}{B - 1}
\sum_{b=1}^{B}
(\theta_b^* - \bar{\theta}^*)^2
}
$$ Where:

-   $B$ : number of bootstrap resamples
-   $\hat{\theta}_b$ : estimator from bootstrap sample $b$
-   $\bar{\hat{\theta}}$ : average bootstrap estimate
:::

------------------------------------------------------------------------

### Bootstrap Confidence Interval

::: {style="font-size:0.4em"}
```{r}

mean_ci <- quantile(
  boot_means,
  probs = c(0.025, 0.975)
)

median_ci <- quantile(
  boot_medians,
  probs = c(0.025, 0.975)
)

tibble(
  estimator = c("Mean TMB", "Median TMB"),
  lower_95_ci = c(mean_ci[1], median_ci[1]),
  upper_95_ci = c(mean_ci[2], median_ci[2])
)
```

**Interpretation**

-   confidence intervals contain the middle 95% of estimates generated from repeated bootstrap resampling

For a 95% bootstrap confidence interval:

-   lower bound is the 2.5th percentile of the bootstrap estimates

-   upper bound is the 97.5th percentile of the bootstrap estimates

-   each bootstrap estimate represents one possible estimate of population TMB obtained from a resampled version of the observed sequencing dataset

-   some bootstrap estimates are smaller and some are larger because sampling variability causes estimates to fluctuate across samples

-   confidence interval summarizes the range of population TMB values that remain plausible given the observed data and the observed variability across resamples
:::

------------------------------------------------------------------------

### Summary

::: {style="font-size:0.4em"}
-   point estimates summarize characteristics of a population using observed sample data

-   different samples produce different estimates because of sampling variability

-   bootstrap resampling approximates repeated sampling directly from the observed dataset

-   bootstrap methods repeatedly:

    -   resample the data with replacement
    -   recompute the estimator
    -   form an empirical sampling distribution

bootstrap distributions can be used to estimate:

-   standard errors
-   confidence intervals
-   estimator stability
:::

------------------------------------------------------------------------

### Summary

::: {style="font-size:0.4em"}
-   point estimates summarize characteristics of a population using observed sample datbootstrap confidence intervals summarize the range of population values that remain plausible given the observed data and the variability observed across bootstrap resamples

-   bootstrap methods are especially useful when:

    -   distributions are skewed
    -   assumptions underlying closed-form methods may be fragile
    -   estimators do not have simple analytic standard error formulas
:::

------------------------------------------------------------------------
