---
title: "Bootstrap Estimation for Mean and Median Biomarker Values"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    chalkboard: true
    smaller: false
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r setup, echo=FALSE}
library(tidyverse)
library(broom)
library(scales)

set.seed(104)

biomarker_df <- tibble(
  id = 1:180,
  treatment = rep(c("Placebo", "Low dose", "High dose"), each = 60),
  ifng = c(
    rlnorm(60, meanlog = log(11.5), sdlog = 0.35),
    rlnorm(60, meanlog = log(15.3), sdlog = 0.38),
    rlnorm(60, meanlog = log(19.8), sdlog = 0.42)
  ),
  response = c(
    rbinom(60, 1, 0.28),
    rbinom(60, 1, 0.45),
    rbinom(60, 1, 0.62)
  ),
  infections = c(
    rpois(60, lambda = 1.4),
    rpois(60, lambda = 1.1),
    rpois(60, lambda = 0.8)
  ),
  age = round(rnorm(180, mean = 61, sd = 10)),
  bmi = round(rnorm(180, mean = 29, sd = 4.5), 1)
) |>
  mutate(
    treatment = factor(treatment, levels = c("Placebo", "Low dose", "High dose")),
    log_ifng = log(ifng)
  )

# add a small number of extreme values to motivate robust estimation
biomarker_extreme_df <- biomarker_df |>
  mutate(
    ifng_extreme = if_else(id %in% c(8, 77, 146), ifng * 4.5, ifng),
    log_ifng_extreme = log(ifng_extreme)
  )
```

### Review

::: {style="font-size:0.63em"}
Previous lectures established what is typically estimated in statistical analysis—defining population quantities and constructing estimators to compare groups \[i.e.: mean differences, risk differences, odds ratios\]

Current lecture focuses on the following:

> Given an estimator and a single observed dataset, how do we quantify its sampling variability when formulas rely on assumptions that may not reflect the observed data?

Bootstrap resampling provides a data-driven way to approximate the sampling distribution of an estimator
:::

------------------------------------------------------------------------

### Lecture Overview

::: {style="font-size:0.63em"}
-   define bootstrap resampling as a method for estimating sampling variability

-   interpret a bootstrap standard error and bootstrap confidence interval

-   explain why robust estimators are useful for skewed or outlier-prone biomedical measurements

-   compare the mean and median as estimators of a typical biomarker value

-   choose an estimation strategy based on the estimand, data structure, assumptions, and scientific question
:::

------------------------------------------------------------------------

### Bootstrap: Definition

::: {style="font-size:0.64em"}
`bootstrap`: an approach that approximates the sampling distribution of an estimator by repeatedly resampling from the observed data with replacement

Bootstrap Approach:

-   treats the observed dataset as a proxy for the population
-   generates many “new” samples by sampling with replacement
-   computes the estimator in each resample
-   uses the resulting distribution to quantify variability
:::

------------------------------------------------------------------------

### Bootstrap: Definition

::: {style="font-size:0.64em"}
Goals within statistical estimation include:

-   how much an estimate varies across samples
-   how precise the estimate is
-   what values are plausible for the population

Approaches rely on:

-   known sampling distributions
-   analytic standard error formulas
-   assumptions about the data \[i.e.: normality\]

However in certain cases: assumptions do not align with observed data

-   closed-form methods for estimating uncertainty rely on assumptions about the data are true in order to estimate sample variability \[i.e.: normality and simple variance structure\]

-   bootstrap methods estimate uncertainty by repeatedly resampling the data and observing how the estimate changes
:::

------------------------------------------------------------------------

### Overview: Bootstrap Procedure

::: {style="font-size:0.64em"}
Given observed data of size n:

**1. Resample with replacement**

-   draw a sample of size n from the original data

**2. Compute the estimator** - i.e.: mean, median, risk difference

**3. Repeat many times** - i.e.: 1,000+ resamples

**4. Form the bootstrap distribution** - collection of all computed estimates

Resulting distribution approximates the sampling distribution of the estimator
:::

------------------------------------------------------------------------

### Overview: Bootstrap Procedure

::: {style="font-size:0.64em"}
From the bootstrap distribution, we can estimate:

1.  **Standard error**: measures variability across resamples

2.  **Confidence intervals**: provides a plausible range for the population parameter

3.  **Shape of the sampling distribution** :

-   symmetry vs skewness
-   sensitivity to outliers
-   stability of the estimator
:::

------------------------------------------------------------------------

### Example Study Background

::: {style="font-size:0.68em"}
Suppose investigators are conducting a clinical study of an **immune-modulating therapy**.

The therapy is designed to alter immune activity, so investigators measure interferon-gamma concentration after treatment.

**Interferon-gamma (IFN-γ)** is a cytokine involved in immune signaling and inflammatory activation.

In this study, IFN-γ is used as a continuous biomarker that may reflect the degree of immune response among participants.

Our study goals are as follows:

Determine:

> What is a typical interferon-gamma (IFN-γ) concentration among patients receiving the immune-modulating therapy?

Because IFN-γ is continuous and often right-skewed:

-   What is the average IFN-γ level in the study population?
-   What is the median IFN-γ level, representing a typical patient?
:::

------------------------------------------------------------------------

### Example Dataset

```{r, echo=FALSE}
library(tidyverse)
library(scales)

set.seed(42)

n <- 180

biomarker_df <- tibble(
  id = 1:n,
  treatment = sample(
    c("Placebo", "Low dose", "High dose"),
    size = n,
    replace = TRUE,
    prob = c(0.34, 0.33, 0.33)
  ),
  age = round(rnorm(n, mean = 58, sd = 11)),
  bmi = round(rnorm(n, mean = 28, sd = 5), 1),
  ifng = rgamma(n, shape = 3.2, scale = 4.5),
  response = rbinom(n, size = 1, prob = 0.48),
  infections = rpois(n, lambda = 1.4)
) |>
  mutate(
    ifng = round(ifng, 2),
    log_ifng = log(ifng)
  )

head(biomarker_df, 5)
```

::: {style="font-size:0.52em"}
**Overview**

-   each row represents one participant
-   `id`: participant identifier
-   `treatment`: placebo, low dose, or high dose
-   `ifng`: interferon-gamma concentration \[continuous and positive\]
-   `response`: whether the participant had a clinical response \[binary\]
-   `infections`: number of infections during follow-up \[count\]
-   `age` and `bmi`: patient characteristics
-   many participants have low-to-moderate values
-   fewer participants have very high values
-   upper tail can pull the mean upward
:::

------------------------------------------------------------------------

### Visualize the Biomarker Distribution

::: {style="font-size:0.55em"}
```{r, echo=FALSE}
plot <- ggplot(biomarker_df, aes(x = ifng)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of IFN-γ Concentration",
    subtitle = "Right-skewed biomarker distributions are common in biomedical data",
    x = "IFN-γ concentration",
    y = "Number of participants"
  ) +
  theme_bw()

plot
```

**Interpretation of the Distribution**

-   histogram shows the distribution of interferon-gamma (IFN-γ) concentrations across study participants
-   distribution is right-skewed \[positively skewed\] $\rightarrow$ because of the skewness the mean is likely larger than the median
:::

------------------------------------------------------------------------

### Visualize the Biomarker Distribution

::: {style="font-size:0.55em"}
```{r, echo=FALSE}
plot <- ggplot(biomarker_df, aes(x = ifng)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of IFN-γ Concentration",
    subtitle = "Right-skewed biomarker distributions are common in biomedical data",
    x = "IFN-γ concentration",
    y = "Number of participants"
  ) +
  theme_bw()

plot
```

**Interpretation of the Distribution**

Note: mean may overstate a typical biomarker level due to extreme values

-   most participants have moderate IFN-γ values \[10–15 pg/ml\]
-   smaller number of participants have substantially higher values, extending toward 40–50
:::

------------------------------------------------------------------------

### Estimation Targets

::: {style="font-size:0.68em"}
We focus on two estimators of a “typical” value:

-   **mean IFN-γ** → average leve \[sensitive to extremes\]\
-   **median IFN-γ** → represents the central point of the distribution \[half the observed immune responses fall below and half fall above\] without being influenced by extreme measurements

Let:

$$
\theta_{mean} = \text{population mean}, \quad
\theta_{median} = \text{population median}
$$

Goal: compute point estimates and quantify uncertainty using two approaches:

-   analytic \[non-bootstrap estimation\]
-   bootstrap estimation
:::

------------------------------------------------------------------------

### Bootstrap Appraoch

::: {style="font-size:0.68em"}
`bootstrap resampling`: approximates repeated sampling using the observed data

Each bootstrap iteration executes the following:

**1. resample IFN-γ values** with replacement

-   after a value is selected, it is returned to the dataset and can be selected again
-   allowing the bootstrap sample to:
    -   maintain same size as the original dataset
    -   note that each resample may include some participants more than once and may omit some participants

**2. compute the estimator on that resampled dataset**

**3. store the result**

**4. repeat many times**

After many iterations the collection of estimates approximates the sampling distribution of the estimator
:::

------------------------------------------------------------------------

### Compute Sample Estimates

::: {style="font-size:0.6em"}
```{r}
mean_ifng <- mean(biomarker_df$ifng)
median_ifng <- median(biomarker_df$ifng)

tibble(
  mean_ifng = mean_ifng,
  median_ifng = median_ifng
)
```

**Interpretation**

-   mean IFN-γ \[14.6 pg/mL\] : average across all participants
-   median IFN-γ \[12.5 pg/mL\] : value that splits the sample in half
-   mean is larger than the median and consistent with right-skewed distribution where higher values pull the average upward

Note each of these values are point estimates:

-   each is a single-number summary computed from this sample
-   each serves as an estimate of a corresponding population quantity
:::

------------------------------------------------------------------------

### Compute Sample Estimates

::: {style="font-size:0.5em"}
```{r}
mean_ifng <- mean(biomarker_df$ifng)
median_ifng <- median(biomarker_df$ifng)

tibble(
  mean_ifng = mean_ifng,
  median_ifng = median_ifng
)
```

**Key Limitations with current values**

Estimates answer:

> What did we observe in this sample?

Estimate values do not answer:

> How much would this estimate change if we repeated the study?

**Motivation for resolving limitations**

-   a different sample of patients could produce different values
-   `sampling variability`: amount of variation across sample
-   without quantifying that variability it is difficult to assess:
    -   how precise these estimates are
    -   how much confidence to place in them
-   Bootstrap methods allow us to estimate how stable these values are across repeated samples
:::

------------------------------------------------------------------------

### Determine Mean Uncertainty

::: {style="font-size:0.66em"}
Compute the following values for the sample mean:

-   standard error \[SE\] : metric that describes variability of the estimator
-   confidence interval : describes plausible population values

$$
SE(\bar{x}) = \frac{s}{\sqrt{n}}
$$

Assumptions:

-   sample is sufficiently large
-   sampling distribution of the mean is approximately normal extreme skewness or outliers do not dominate the sample
:::

------------------------------------------------------------------------

### Determine Mean Uncertainty

::: {style="font-size:0.66em"}
```{r}
mean_se <- sd(biomarker_df$ifng) / sqrt(nrow(biomarker_df))
mean_ci <- mean_ifng + c(-1.96, 1.96) * mean_se

tibble(
  mean = mean_ifng,
  se = mean_se,
  lower = mean_ci[1],
  upper = mean_ci[2]
)
```

**Interpretation**

-   population average IFN-γ: 14.6 pg/mL
-   true population mean IFN-γ lies between 13.3 and 15.8 pg/mL
-   standard error: 0.622
    -   if we repeated the study many times $\rightarrow$ mean would vary by about 0.6 units
:::

------------------------------------------------------------------------

### Estimating Median

::: {style="font-size:0.68em"}
-   median is often an ideal metric for skewed data
-   metirc represents a typical patient without being influenced by extreme values.
-   median is not defined by magnitude
-   depends on the ordering of the data
-   not the numerical contribution of each value
-   small changes near the center can shift the median
-   large changes in extreme values may have no effect
:::

------------------------------------------------------------------------

### Median Estimation

::: {style="font-size:0.68em"}
```{r, echo=FALSE}
location_summary <- biomarker_df |>
  summarise(
    # Core estimator
    median_ifng = median(ifng),
    
    # Spread around the median
    iqr_ifng = IQR(ifng),
    
    # Key quantiles for interpretation
    q25 = quantile(ifng, 0.25),
    q75 = quantile(ifng, 0.75),
    q90 = quantile(ifng, 0.90)
  )

location_summary
```

**Interpretation**

**Median**

-   median IFN-γ : 15.62 pg/mL

**Interquartile Range \[IQR\]**

IQR: 7.99 pg/mL

-   width of that middle range
-   indicates a moderate spread around the typical value
:::

------------------------------------------------------------------------

### Median Estimation

::: {style="font-size:0.68em"}
```{r, echo=FALSE}
location_summary <- biomarker_df |>
  summarise(
    # Core estimator
    median_ifng = median(ifng),
    
    # Spread around the median
    iqr_ifng = IQR(ifng),
    
    # Key quantiles for interpretation
    q25 = quantile(ifng, 0.25),
    q75 = quantile(ifng, 0.75),
    q90 = quantile(ifng, 0.90)
  )

location_summary
```

**Interpretation**

**q90**: describes upper tail of the distribution or unusually high IFN-γ responses

Q90 = 23.27 pg/mL

-   90% of participants are below this value
-   top 10% of patients have values exceed \~23 pg/mL

**First \[Q1\] and Third Quartile**

Q25 = 11.52 pg/mL Q75 = 19.51 pg/mL

-   middle 50% of participants have IFN-γ values between \~11.5 and \~19.5 pg/mL
:::

------------------------------------------------------------------------

### Distribution Comparison: Closed-Form vs Bootstrap

::: {style="font-size:0.64em"}
```{r, echo=FALSE}
library(tidyverse)

# ---------------------------------------
# Setup
# ---------------------------------------

mean_ifng <- mean(biomarker_df$ifng)
n <- nrow(biomarker_df)
sd_ifng <- sd(biomarker_df$ifng)
se_closed <- sd_ifng / sqrt(n)

# ---------------------------------------
# Bootstrap distribution (empirical)
# ---------------------------------------

set.seed(901)

boot_means <- replicate(2000, {
  x_star <- sample(
    biomarker_df$ifng,
    size = n,
    replace = TRUE
  )
  mean(x_star)
})

boot_df <- tibble(
  value = boot_means,
  method = "Bootstrap (empirical)"
)

# ---------------------------------------
# Closed-form distribution (theoretical)
# ---------------------------------------

theory_means <- rnorm(
  2000,
  mean = mean_ifng,
  sd = se_closed
)

theory_df <- tibble(
  value = theory_means,
  method = "Closed-form \n (normal approximation)"
)

# ---------------------------------------
# Combine and plot
# ---------------------------------------

plot_df <- bind_rows(boot_df, theory_df)

plot <- ggplot(plot_df, aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~method, scales = "free_y") +
  geom_vline(
    xintercept = mean_ifng,
    linetype = "dashed"
  ) +
  labs(
    title = "Sampling Distribution of the Mean: Closed-Form vs Bootstrap",
    subtitle = "Both represent how the sample mean varies across repeated samples",
    x = "Sample Mean IFN-γ (pg/mL)",
    y = "Count"
  ) +
  theme_bw()
plot
```

**Objective(s)**:

-   compare two approaches for approximating IFN-γ population mean
    -   evaluate whether a closed form approximation and a data-driven method \[bootstrap\] produce similar representations of uncertainty
:::

------------------------------------------------------------------------

### Distribution Comparison: Closed-Form vs Bootstrap

```{r, echo=FALSE}

#| fig-width: 4
#| fig-height: 3
#| 
library(tidyverse)

# ---------------------------------------
# Setup
# ---------------------------------------

mean_ifng <- mean(biomarker_df$ifng)
n <- nrow(biomarker_df)
sd_ifng <- sd(biomarker_df$ifng)
se_closed <- sd_ifng / sqrt(n)

# ---------------------------------------
# Bootstrap distribution (empirical)
# ---------------------------------------

set.seed(901)

boot_means <- replicate(2000, {
  x_star <- sample(
    biomarker_df$ifng,
    size = n,
    replace = TRUE
  )
  mean(x_star)
})

boot_df <- tibble(
  value = boot_means,
  method = "Bootstrap (empirical)"
)

# ---------------------------------------
# Closed-form distribution (theoretical)
# ---------------------------------------

theory_means <- rnorm(
  2000,
  mean = mean_ifng,
  sd = se_closed
)

theory_df <- tibble(
  value = theory_means,
  method = "Closed-form \n (normal approximation)"
)

# ---------------------------------------
# Combine and plot
# ---------------------------------------

plot_df <- bind_rows(boot_df, theory_df)

plot <- ggplot(plot_df, aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~method, scales = "free_y") +
  geom_vline(
    xintercept = mean_ifng,
    linetype = "dashed"
  ) +
  labs(
    title = "Sampling Distribution of the Mean: Closed-Form vs Bootstrap",
    subtitle = "Both represent how the sample mean varies across repeated samples",
    x = "Sample Mean IFN-γ (pg/mL)",
    y = "Count"
  ) +
  theme_bw()

```

::: {style="font-size:0.55em"}
```{r}
#| fig-width: 4
#| fig-height: 3

plot
```

**Interpretation**

**Bootstrap Panel**

-   2000 resampled datasets drawn from the observed data \[n = 180 patients\]

**Closed-form panel**:

-   2000 simulated values drawn from a theoretical distribution
-   each value represents a possible sample mean
:::

------------------------------------------------------------------------

### Distribution Comparison: Closed-Form vs Bootstrap

```{r, echo=FALSE}

#| fig-width: 4
#| fig-height: 3
#| 
library(tidyverse)

# ---------------------------------------
# Setup
# ---------------------------------------

mean_ifng <- mean(biomarker_df$ifng)
n <- nrow(biomarker_df)
sd_ifng <- sd(biomarker_df$ifng)
se_closed <- sd_ifng / sqrt(n)

# ---------------------------------------
# Bootstrap distribution (empirical)
# ---------------------------------------

set.seed(901)

boot_means <- replicate(2000, {
  x_star <- sample(
    biomarker_df$ifng,
    size = n,
    replace = TRUE
  )
  mean(x_star)
})

boot_df <- tibble(
  value = boot_means,
  method = "Bootstrap (empirical)"
)

# ---------------------------------------
# Closed-form distribution (theoretical)
# ---------------------------------------

theory_means <- rnorm(
  2000,
  mean = mean_ifng,
  sd = se_closed
)

theory_df <- tibble(
  value = theory_means,
  method = "Closed-form \n (normal approximation)"
)

# ---------------------------------------
# Combine and plot
# ---------------------------------------

plot_df <- bind_rows(boot_df, theory_df)

plot <- ggplot(plot_df, aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~method, scales = "free_y") +
  geom_vline(
    xintercept = mean_ifng,
    linetype = "dashed"
  ) +
  labs(
    title = "Sampling Distribution of the Mean: Closed-Form vs Bootstrap",
    subtitle = "Both represent how the sample mean varies across repeated samples",
    x = "Sample Mean IFN-γ (pg/mL)",
    y = "Count"
  ) +
  theme_bw()

```

::: {style="font-size:0.55em"}
```{r}
#| fig-width: 4
#| fig-height: 3

plot
```

**Comparison**

-   each histogram represents \~2000 possible sample means
-   dashed line marks the observed sample mean
-   both panels show \~2000 possible values of the sample mean
-   bootstrap generates values by resampling full datasets
-   closed-form approach generates values directly from a theoretical normal distribution
-   both bootstrap and closed-form methods produce similar sampling distributions for the mean
-   indicates that bootstrap based estimate is stable across repeated samples
:::

------------------------------------------------------------------------

### Summary: Bootstrap and Robust Estimation

::: {style="font-size:0.64em"}
Bootstrap and robust methods solve related but distinct problems.

**Bootstrap**

-   estimates uncertainty
-   approximates the sampling distribution
-   supports standard errors and intervals

**Robust estimation**

-   improves stability
-   reduces sensitivity to extreme values
-   helps summarize skewed biomedical data

Together, they strengthen estimation when classical assumptions are fragile.
:::

------------------------------------------------------------------------

### Summary

::: {style="font-size:0.66em"}
-   estimates vary across repeated samples

-   point estimates summarize values from a single dataset\

-   sampling variability determines how estimates change

In practice:

-   different samples produce different estimates\

-   stability of an estimate depends on data structure and sample size

-   **bootstrap methods**: resample observed data to approximate the sampling distribution

-   multiple approaches can yield similar conclusions\

-   agreement between methods supports stability of the estimate
:::
